---
title: "Comparing Indices"
output:
  html_document: default
  html_notebook: default
---
####Note for user regarding this R Notebook:
This notebook compares all indices to one another. In order to knit the document fully, we must first include all the code to build the indices (including loading and subsetting the data, re-leveling the variables, and building the MCA models), as the comparison plots reference the values generated by the MCA models. These pieces of code will make the notebook longer to knit, but will not appear in the output file.


####Pulling/manipulating data before importing to R
In IPUMS-I: Select variables to include as many of  
&nbsp;&nbsp;&nbsp;&nbsp;(COUNTRY, YEAR, SAMPLE, SERIAL, PERNUM, PERWT, RESIDENT preselected),  
&nbsp;&nbsp;&nbsp;&nbsp;HHWT, PERSONS, OWNERSHIP, ELECTRIC, WATSUP, SEWAGE, FUELCOOK, TRASH, HEAT, AIRCON, ROOMS, TOILET, FLOOR, WALL, ROOF,  
&nbsp;&nbsp;&nbsp;&nbsp;PHONE, CELL, AUTOS, REFRIG, TV, RADIO,  
&nbsp;&nbsp;&nbsp;&nbsp;CHBORN, SCHOOL, EDATTAIN, AND OCCISCO  
as are available.

In JMP: Uncheck "Use value labels" for all variables that were not preselected.

Note: Search for USER ACTION to find places in this code that require user consideration for each country sample. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(magrittr)
library(FactoMineR)
library(ggplot2)
```

```{r data source, include = FALSE}
#USER ACTION: replace "Zambia_2010.csv" with data from country of interest
#country <- read.csv("<csv filename here>")
country <- read.csv("Zambia_2010.csv")
```

```{r subsetting detailed, include = FALSE}
#Index 1
#USER ACTION: Comment out lines of code for variables unavailable in country sample:
country_1 = country %>% 
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier, 
  Ownership.of.dwelling..general.version., 
  Person.number, 
  Residence.status..de.facto..de.jure, 
  Children.ever.born, 
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general, 
  Telephone.availability,
  Cellular.phone.availability, 
  Automobiles.available, 
  Refrigerator, 
  Television.set, 
  Radio.in.household
  ))

#Index 2
country_2 = country %>% 
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier, 
  Ownership.of.dwelling..detailed.version., 
  Person.number, 
  Residence.status..de.facto..de.jure, 
  Children.ever.born, 
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general, 
  Trash.disposal,
  #Central.heating,
  #Air.conditioning,
  Telephone.availability,
  Cellular.phone.availability, 
  Automobiles.available, 
  Refrigerator, 
  Television.set, 
  Radio.in.household
  ))

#Index 3
country_3 = country %>%
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier,
  Ownership.of.dwelling..general.version.,
  Person.number,
  Residence.status..de.facto..de.jure,
  Children.ever.born,
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general
))

#Index 4
country_4 = country %>%
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier,
  Ownership.of.dwelling..detailed.version.,
  Person.number,
  Residence.status..de.facto..de.jure,
  Children.ever.born,
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general,
  #Central.heating,
  #Air.conditioning,
  Trash.disposal
))
```

```{r function not in, include = FALSE}
#Make "Not in" function (used for all indices)
'%!in%' <- function(x,y)!('%in%'(x,y))
```

```{r, include = FALSE}
# USER ACTION: Check this code for each data set. Comment out lines of code for variables unavailable in country sample.  
# Note: Pay special attention to SEWAGE (see notes below) - needs to be checked for each sample  
# Note: As written, this code removes NIU, Unknown for each variable. User may remove or change filter commands to include NIU, Unknown.
```

```{r releveling index1, include = FALSE}
#Making variables for Index 1

#Making ownership.1 (Owned, Rented, or Other ownership status)
country_1 <- country_1 %>% filter(Ownership.of.dwelling..detailed.version. %!in% c(0,999)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..detailed.version. %in% 100:194, "Owned", ifelse(Ownership.of.dwelling..detailed.version. %in% 210:239, "Rented", "Other ownership status")))

#Making electric.1 (Has electricity or No electricity)
country_1 <- country_1 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply.1 (Access to piped water or No access to piped water)
country_1 <- country_1 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Note: Sewage is a variable that needs manual attention for Indices 1 and 3. Either, Sewage is ready to be simplified for Index 1 or Sewage is in the form intended for Index 2. In the second case, Sewage may be excluded or included (in simplified form) in Index 1 (at this stage of the research). A more formal or educated recommendation may follow in the course of our research.
# Run <unique(country_1$Sewage)> to check form of SEWAGE for sample.

#Making sewage.simple.1 (Sewage access [septic or public] or No sewage disposal)
country_1 <- country_1 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.simple.1 = ifelse(Sewage == 20, "No sewage disposal", "Sewage access"))

# #Making sewage.detailed.1 (Sewage system, Septic system, or No sewage)
# country_1 <- country_1 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.detailed.1 = ifelse(Sewage == 11, "Sewage system", ifelse(Sewage == 12, "Septic system", "No sewage")))

#Making cooking.fuel.1 (Cook with electricity, Cook with fossil fuel, Cook with solid fuel, or Cook with other fuel)
country_1 <- country_1 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel == 20, "Cook with electricity", ifelse(Cooking.fuel %in% c(30:47,62:65), "Cook with fossil fuel", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel"))))

#Making trash.1 (Collected by service or No sanitation service)
country_1 <- country_1 %>% filter(Trash.disposal %!in% c(0,99)) %>% mutate(trash.1 = ifelse(Trash.disposal %in% 10:14, "Trash collected by service", "No sanitation service"))

# #Making heat.1 (Central heating or No central heating)
# country_1 <- country_1 %>% filter(Central.heating %!in% c(0,9)) %>% mutate(heat.1 = ifelse(Central.heating %in% c(1, 5:7), "No central heating", "Central heating"))
# 
# #Alternative: If available for country sample, make aircon.1 to fill in for heat (No air conditioning or Air conditioning)
# country_1 <- country_1 %>% filter(Air.conditioning %!in% c(0,99)) %>% mutate(aircon.1 = ifelse(Air.conditioning == 10, "No air conditioning", "Air conditioning"))

#Note: We kept Number.of.rooms as an integer, assuming that making all households with code 30 (30+ rooms) equate to 30 rooms would have a negligable effect on the index; very few houses will have greater than 30 rooms, and households with enough wealth for 30 rooms will be ranked similarly to those with greater than 30.
#Note: We changed Number.of.rooms code 00 (Part of a room; no rooms) to 1 in order to do the calculation (Number.of.person.records/Number.of.rooms), assuming that having part of a room or no room is similar enough to having only one room to have a negligable effect.
#Note: We rounded number.ppl.per.room.1 to limit the levels that would result in the MCA to only integer values instead of all possible numeric values
#Note: We also tried binning number.ppl.per.room.1 to be in ranges: 0-2, 3-9, 10+ people per room, as the initial MCA gave disproportional weight to the levels of number.ppl.per.room.1.

#Making number.ppl.per.room.1
country_1 <- country_1 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_1 <- country_1 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/number.of.rooms.1)
country_1$number.ppl.per.room.1 <- round(country_1$number.ppl.per.room.1)
country_1$number.ppl.per.room.1 <- as.integer(country_1$number.ppl.per.room.1) 

#Different cutoffs for binning: Making bin.number.ppl.per.room.1
country_1 <- country_1 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_1 <- country_1 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Flush toilet, Non-flush toilet, No toilet)
country_1 <- country_1 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet == 10, "No toilet",ifelse(Toilet == 21, "Flush toilet","Non-flush toilet")))

#Making floor.1 (Finished or Unfinished)
country_1 <- country_1 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), "Finished floors", "Unfinished floors"))

#Making wall.1 (Cement or Not cement)
country_1 <- country_1 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), "Cement walls", "Not cement walls"))

#Making roof.1 (Cement or Not cement)
country_1 <- country_1 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), "Cement roof", "Not cement roof"))
```

```{r releveling index2, include = FALSE}
#Making variables for Index 2

#Making ownership.1 (Owned or Not owned)
country_2 <- country_2 %>% filter(Ownership.of.dwelling..general.version. %!in% c(0,9)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..general.version. == 1, "Owned","Not owned"))

#Making electric.1 (Has electricity or No electricity)
country_2 <- country_2 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply.1 (Access to piped water or No access to piped water)
country_2 <- country_2 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Making sewage.general.1 (Sewage system or No sewage system)
country_2 <- country_2 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.general.1 = ifelse(Sewage %in% 10:12, "Sewage system", "No sewage system"))

#Making cooking.fuel.1 (Cook with fossil fuel or elec, Cook with solid fuels, or Cook with other fuel)
country_2 <- country_2 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel %in% c(20,30:47,62:65), "Cook with fossil fuel or elec", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel")))

#See Index 1 number.ppl.per.room.1 for notes regarding this calculation
#Making number.ppl.per.room.1
country_2 <- country_2 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(Number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_2 <- country_2 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/Number.of.rooms.1)
country_2$number.ppl.per.room.1 <- round(country_2$number.ppl.per.room.1)
country_2$number.ppl.per.room.1 <- as.integer(country_2$number.ppl.per.room.1) 

#Different binning: Making bin.number.ppl.per.room.1
country_2 <- country_2 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_2 <- country_2 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Toilet or No toilet)
country_2 <- country_2 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet %in% 10:11, "No toilet","Toilet"))

#Making dwelling.1 (Has cement structures [cement floor, cement wall, or cement roof] or has No cement structures) - need to create each of floor.1, wall.1, roof.1 first
#floor.1 (1 [Finished] or 0 [Unfinished])
country_2 <- country_2 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), 1, 0))

#wall.1 (1 [Cement] or 0 [Not cement])
country_2 <- country_2 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), 1, 0))

#roof.1 (1 [Cement] or 0 [Not cement])
country_2 <- country_2 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), 1, 0))

#floor.wall.roof.1 (sum of floor.1, wall.1, roof.1)
country_2 <- country_2 %>% mutate(floor.wall.roof.1 = floor.1+wall.1+roof.1)
#dwelling.1
country_2 <- country_2 %>% mutate(dwelling.1 = ifelse(floor.wall.roof.1 > 0, "Cement structures", "No cement structures"))
```

```{r releveling index3, include = FALSE}
#Making variables for Index 3

#Making ownership.1 (Owned, Rented, or Other)
country_3 <- country_3 %>% filter(Ownership.of.dwelling..detailed.version. %!in% c(0,999)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..detailed.version. %in% 100:194, "Owned", ifelse(Ownership.of.dwelling..detailed.version. %in% 210:239, "Rented", "Other ownership status")))

#Making electric.1 (Has electricity or No electricity)
country_3 <- country_3 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply1 (Access to piped water or No access to piped water)
country_3 <- country_3 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Note: See Index 1 sewage.1 for notes regarding this variable
#Run <unique(country_3$Sewage)> to check form of SEWAGE for sample.

#Making sewage.simple.1 (Sewage access [septic or public] or No sewage disposal)
country_3 <- country_3 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.simple.1 = ifelse(
    Sewage == 20, "No sewage disposal", "Sewage access"))

# #Making sewage.detailed.1 (Sewage system, Septic system, or No sewage)
# country_3 <- country_3 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.detailed.1 = ifelse(Sewage == 11, "Sewage system", ifelse(Sewage == 12, "Septic system", "No sewage")))

#Making cooking.fuel.1 (Cook with electricity, Cook with fossil fuel, Cook with solid fuel, or Cook with other fuel)
country_3 <- country_3 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel == 20, "Cook with electricity", ifelse(Cooking.fuel %in% c(30:47,62:65), "Cook with fossil fuel", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel"))))

#Making trash.1 (Collected by service or No sanitation service)
country_3 <- country_3 %>% filter(Trash.disposal %!in% c(0,99)) %>% mutate(trash.1 = ifelse(Trash.disposal %in% 10:14, "Trash collected by service", "No sanitation service"))

# #Making heat.1 (Central heating or No central heating)
# country_3 <- country_3 %>% filter(Central.heating %!in% c(0,9)) %>% mutate(heat.1 = ifelse(Central.heating %in% c(1, 5:7), "No central heating", "Central heating"))
#
# #Alternative: Make aircon.1 to fill in for heat (No air conditioning or Air conditioning)
# country_3 <- country_3 %>% filter(Air.conditioning %!in% c(0,99)) %>% mutate(aircon.1 = ifelse(Air.conditioning == 10, "No air conditioning", "Air conditioning"))

#Note: See Index 1 number.ppl.per.room.1 for notes regarding this calculation
#Making number.ppl.per.room.1
country_3 <- country_3 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_3 <- country_3 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/number.of.rooms.1)
country_3$number.ppl.per.room.1 <- round(country_3$number.ppl.per.room.1)
country_3$number.ppl.per.room.1 <- as.integer(country_3$number.ppl.per.room.1)

#Different binning: Making bin.number.ppl.per.room.1
country_3 <- country_3 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_3 <- country_3 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Flush toilet, Non-flush toilet, or No toilet)
country_3 <- country_3 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet == 10, "No toilet",ifelse(Toilet == 21, "Flush toilet","Non-flush toilet")))

#Making floor.1 (Finished or Unfinished)
country_3 <- country_3 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), "Finished floors", "Unfinished floors"))

#Making wall.1 (Cement or Not cement)
country_3 <- country_3 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), "Cement walls", "Not cement walls"))

#Making roof.1 (Cement or Not cement)
country_3 <- country_3 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), "Cement roof", "Not cement roof"))

#Adding appliances:

#For phone, cell: Make telephone.1 and cell.1, and telephone.cell.1 (column indicating whether household has at least one of phone.1 or cell.1)
#Making telephone.1 (1 [Telephone] or 0 [no telephone]); ALSO making cell.1 (1 [cell] or 0 [no cell])
country_3 <- country_3 %>% filter(Telephone.availability %!in% c(0,9) & Cellular.phone.availability %!in% c(0,9)) %>%
  mutate(telephone.1 = ifelse(Telephone.availability == 2, 1, 0),
         cell.1 = ifelse(Cellular.phone.availability == 1, 1, 0))
#Making telephone.cell.1
country_3 <- country_3 %>% mutate(telephone.cell.sum = telephone.1 + cell.1)
country_3 <- country_3 %>% mutate(telephone.cell.1 = ifelse(telephone.cell.sum == 0, "No phone or cell", "Phone or cell"))

#Making autos.1 (Have auto or No auto)
country_3 <- country_3 %>% filter(Automobiles.available %!in% c(8,9)) %>% mutate(autos.1 = ifelse(Automobiles.available == 0, "No auto", "Have auto"))

#Making refrig.1 (Have refrig or No refrig)
country_3 <- country_3 %>% filter(Refrigerator %!in% c(0,9)) %>% mutate(refrig.1 = ifelse(Refrigerator == 2, "Have refrig","No refrig"))

#Making TV.1 (Have TV or No TV)
country_3 <- country_3 %>% filter(Television.set %!in% c(0,99)) %>% mutate(TV.1 = ifelse(Television.set == 10, "No TV", "Have TV"))

#Making radio.1 (Have radio or No radio)
country_3 <- country_3 %>% filter(Radio.in.household %!in% c(0,9)) %>% mutate(radio.1 = ifelse(Radio.in.household == 2, "Have radio","No radio"))
```

```{r releveling index4, include = FALSE}
#Making variables for Index 4

#Making ownership.1 (Owned or Not owned)
country_4 <- country_4 %>% filter(Ownership.of.dwelling..general.version. %!in% c(0,9)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..general.version. == 1, "Owned","Not owned"))

#Making electric.1 (Has electricity or No electricity)
country_4 <- country_4 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply.1 (Access to piped water or No access to piped water)
country_4 <- country_4 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Making sewage.general.1 (Sewage system or No sewage system)
country_4 <- country_4 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.general.1 = ifelse(Sewage %in% 10:12, "Sewage system", "No sewage system"))

#Making cooking.fuel.1 (Cook with fossil fuel or elec, Cook with solid fuels, or Cook with other fuel)
country_4 <- country_4 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel %in% c(20,30:47,62:65), "Cook with fossil fuel or elec", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel")))

#See Index 1 number.ppl.per.room.1 for notes regarding this calculation
#Making number.ppl.per.room.1
country_4 <- country_4 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(Number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_4 <- country_4 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/Number.of.rooms.1)
country_4$number.ppl.per.room.1 <- round(country_4$number.ppl.per.room.1)
country_4$number.ppl.per.room.1 <- as.integer(country_4$number.ppl.per.room.1) 

#Different binning: Making bin.number.ppl.per.room.1
country_4 <- country_4 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_4 <- country_4 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Toilet or No toilet)
country_4 <- country_4 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet %in% 10:11, "No toilet","Toilet"))

#Making dwelling.1 (Has cement structures [cement floor, cement wall, or cement roof] or has No cement structures) - need to create each of floor.1, wall.1, roof.1 first
#floor.1 (1 [Finished] or 0 [Unfinished])
country_4 <- country_4 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), 1, 0))

#wall.1 (1 [Cement] or 0 [Not cement])
country_4 <- country_4 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), 1, 0))

#roof.1 (1 [Cement] or 0 [Not cement])
country_4 <- country_4 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), 1, 0))

#floor.wall.roof.1 (sum of floor.1, wall.1, roof.1)
country_4 <- country_4 %>% mutate(floor.wall.roof.1 = floor.1+wall.1+roof.1)
#dwelling.1
country_4 <- country_4 %>% mutate(dwelling.1 = ifelse(floor.wall.roof.1 > 0, "Cement structures", "No cement structures"))

#Adding appliances:

#For phone, cell: Make telephone.1 and cell.1, and telephone.cell.1 (column indicating whether household has at least one of phone.1 or cell.1)
#Making telephone.1 (1 [Telephone] or 0 [no telephone]); ALSO making cell.1 (1 [cell] or 0 [no cell])
country_4 <- country_4 %>% filter(Telephone.availability %!in% c(0,9) & Cellular.phone.availability %!in% c(0,9)) %>%
  mutate(telephone.1 = ifelse(Telephone.availability == 2, 1, 0),
         cell.1 = ifelse(Cellular.phone.availability == 1, 1, 0))
#Making telephone.cell.1
country_4 <- country_4 %>% mutate(telephone.cell.sum = telephone.1 + cell.1)
country_4 <- country_4 %>% mutate(telephone.cell.1 = ifelse(telephone.cell.sum == 0, "No phone or cell", "Phone or cell"))

#Making autos.1 (Have auto or No radio)
country_4 <- country_4 %>% filter(Automobiles.available %!in% c(8,9)) %>% mutate(autos.1 = ifelse(Automobiles.available == 0, "No auto", "Have auto"))

#Making refrig.1 (Have refrig or No radio)
country_4 <- country_4 %>% filter(Refrigerator %!in% c(0,9)) %>% mutate(refrig.1 = ifelse(Refrigerator == 2, "Have refrig","No refrig"))

#Making TV.1 (Have TV or No radio)
country_4 <- country_4 %>% filter(Television.set %!in% c(0,99)) %>% mutate(TV.1 = ifelse(Television.set == 10, "No TV", "Have TV"))

#Making radio.1 (Have radio or No radio)
country_4 <- country_4 %>% filter(Radio.in.household %!in% c(0,9)) %>% mutate(radio.1 = ifelse(Radio.in.household == 2, "Have radio","No radio"))
```

```{r change to factors, include = FALSE}
# USER ACTION:
# Note: Comment out lines of code for variables unavailable in country sample.  
# Note: Review Sewage variable choice before running this chunk. 

#Index 1
country_1$ownership.1 = as.factor(country_1$ownership.1)
country_1$electric.1 = as.factor(country_1$electric.1)
country_1$water.supply.1 = as.factor(country_1$water.supply.1)

#Sewage: Comment out the irrelevant variable (sewage.<simple or detailed>.1)
country_1$sewage.simple.1 = as.factor(country_1$sewage.simple.1)
#country_1$sewage.detailed.1 = as.factor(country_1$sewage.detailed.1)

country_1$cooking.fuel.1 = as.factor(country_1$cooking.fuel.1)
country_1$trash.1 = as.factor(country_1$trash.1)
#country_1$heat.1 = as.factor(country_1$heat.1)
#country_1$aircon.1 = as.factor(country_1$aircon.1)
country_1$number.ppl.per.room.1 = as.factor(country_1$number.ppl.per.room.1)
country_1$bin.number.ppl.per.room.1 = as.factor(country_1$bin.number.ppl.per.room.1)
country_1$alt.bin.number.ppl.per.room.1 = as.factor(country_1$alt.bin.number.ppl.per.room.1)
country_1$toilet.1 = as.factor(country_1$toilet.1)
country_1$floor.1 = as.factor(country_1$floor.1)
country_1$wall.1 = as.factor(country_1$wall.1)
country_1$roof.1 = as.factor(country_1$roof.1)


#Index 2
country_2$ownership.1 = as.factor(country_2$ownership.1)
country_2$electric.1 = as.factor(country_2$electric.1)
country_2$water.supply.1 = as.factor(country_2$water.supply.1)
country_2$sewage.general.1 = as.factor(country_2$sewage.general.1)
country_2$cooking.fuel.1 = as.factor(country_2$cooking.fuel.1)
country_2$number.ppl.per.room.1 = as.factor(country_2$number.ppl.per.room.1)
country_2$bin.number.ppl.per.room.1 = as.factor(country_2$bin.number.ppl.per.room.1)
country_2$alt.bin.number.ppl.per.room.1 = as.factor(country_2$alt.bin.number.ppl.per.room.1)
country_2$toilet.1 = as.factor(country_2$toilet.1)
country_2$dwelling.1 = as.factor(country_2$dwelling.1)


#Index 3
country_3$ownership.1 = as.factor(country_3$ownership.1)
country_3$electric.1 = as.factor(country_3$electric.1)
country_3$water.supply.1 = as.factor(country_3$water.supply.1)

#Sewage: Comment out the irrelevant variable (sewage.<simple or detailed>.1)
country_3$sewage.simple.1 = as.factor(country_3$sewage.simple.1)
#country_3$sewage.detailed.1 = as.factor(country_3$sewage.detailed.1)

country_3$cooking.fuel.1 = as.factor(country_3$cooking.fuel.1)
country_3$trash.1 = as.factor(country_3$trash.1)
#country_3$heat.1 = as.factor(country_3$heat.1)
#country_3$aircon.1 = as.factor(country_3$aircon.1)
country_3$number.ppl.per.room.1 = as.factor(country_3$number.ppl.per.room.1)
country_3$bin.number.ppl.per.room.1 = as.factor(country_3$bin.number.ppl.per.room.1)
country_3$alt.bin.number.ppl.per.room.1 = as.factor(country_3$alt.bin.number.ppl.per.room.1)
country_3$toilet.1 = as.factor(country_3$toilet.1)
country_3$floor.1 = as.factor(country_3$floor.1)
country_3$wall.1 = as.factor(country_3$wall.1)
country_3$roof.1 = as.factor(country_3$roof.1)
country_3$telephone.cell.1 = as.factor(country_3$telephone.cell.1)
country_3$autos.1 = as.factor(country_3$autos.1)
country_3$refrig.1 = as.factor(country_3$refrig.1)
country_3$TV.1 = as.factor(country_3$TV.1)
country_3$radio.1 = as.factor(country_3$radio.1)


#Index 4
country_4$ownership.1 = as.factor(country_4$ownership.1)
country_4$electric.1 = as.factor(country_4$electric.1)
country_4$water.supply.1 = as.factor(country_4$water.supply.1)
country_4$sewage.general.1 = as.factor(country_4$sewage.general.1)
country_4$cooking.fuel.1 = as.factor(country_4$cooking.fuel.1)
country_4$number.ppl.per.room.1 = as.factor(country_4$number.ppl.per.room.1)
country_4$bin.number.ppl.per.room.1 = as.factor(country_4$bin.number.ppl.per.room.1)
country_4$alt.bin.number.ppl.per.room.1 = as.factor(country_4$alt.bin.number.ppl.per.room.1)
country_4$toilet.1 = as.factor(country_4$toilet.1)
country_4$dwelling.1 = as.factor(country_4$dwelling.1)
country_4$telephone.cell.1 = as.factor(country_4$telephone.cell.1)
country_4$autos.1 = as.factor(country_4$autos.1)
country_4$refrig.1 = as.factor(country_4$refrig.1)
country_4$TV.1 = as.factor(country_4$TV.1)
country_4$radio.1 = as.factor(country_4$radio.1)
```

```{r serial numbers, include = FALSE}
# USER ACTION: If using a sample containing more than one country, the row name needs to contain the full IPUMS idendifier and Household.serial.number to track the individual houses through the full MCA analysis. 

row.names(country_1) = country_1$Household.serial.number
row.names(country_2) = country_2$Household.serial.number
row.names(country_3) = country_3$Household.serial.number
row.names(country_4) = country_4$Household.serial.number
```

```{r mca1234, include = FALSE}
#USER ACTION: Comment out variables unavailable in the country sample:

country_1.sub = select(country_1, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.simple.1, 
                                      #sewage.detailed.1,
                                      cooking.fuel.1,
                                      trash.1,
                                      #heat.1,
                                      #aircon.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      floor.1,
                                      wall.1,
                                      roof.1))

country_2.sub = select(country_2, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.general.1,
                                      cooking.fuel.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      dwelling.1))

country_3.sub = select(country_3, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.simple.1, 
                                      #sewage.detailed.1,
                                      cooking.fuel.1,
                                      trash.1,
                                      #heat.1,
                                      #aircon.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      floor.1,
                                      wall.1,
                                      roof.1,
                                      telephone.cell.1,
                                      autos.1,
                                      refrig.1,
                                      TV.1,
                                      radio.1))

country_4.sub = select(country_4, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.general.1,
                                      cooking.fuel.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      dwelling.1,
                                      telephone.cell.1,
                                      autos.1,
                                      refrig.1,
                                      TV.1,
                                      radio.1))

country_1.mca = MCA(country_1.sub,row.w = country_1$Household.weight, graph = FALSE)
country_2.mca = MCA(country_2.sub,row.w = country_2$Household.weight, graph = FALSE)
country_3.mca = MCA(country_3.sub,row.w = country_3$Household.weight, graph = FALSE)
country_4.mca = MCA(country_4.sub,row.w = country_4$Household.weight, graph = FALSE)
```

```{r household score, include = FALSE}
#Index 1
households_1 <- data.frame(country_1.mca$ind$coord[,1])
households_1$ID <- rownames(households_1)
colnames(households_1) <- c("score","ID")

#Index 2
households_2 <- data.frame(country_2.mca$ind$coord[,1])
households_2$ID <- rownames(households_2)
colnames(households_2) <- c("score","ID")

#Index 3
households_3 <- data.frame(country_3.mca$ind$coord[,1])
households_3$ID <- rownames(households_3)
colnames(households_3) <- c("score","ID")

#Index 4
households_4 <- data.frame(country_4.mca$ind$coord[,1])
households_4$ID <- rownames(households_4)
colnames(households_4) <- c("score","ID")
```

##Index Comparisons
USER ACTION: Begin reviewing code for index comparisons here.  

####Scatterplots: 
#####Household wealth scores of Index 2 vs Indicies 1, 3, and 4

Index 2 vs Index 1:
```{r scatter 2vs1, echo = FALSE}
households_2_1 <- inner_join(households_2,households_1,by="ID")
colnames(households_2_1) <- c("Index2","ID","Index1")

ggplot(data = households_2_1) + 
  geom_point(aes(x = Index2,y = Index1), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_smooth(aes(x = Index2,y = Index1),
              color = "red2", method = "lm") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Household Wealth Scores") +
  xlab("Index 2") +
  ylab("Index 1") +
  xlim(-0.8,2.15) +
  ylim(-0.8,2.15) +
  coord_equal(ratio = 1) 
```

Index 2 vs Index 3:  
```{r scatter 2vs3, echo = FALSE}
households_2_3 <- inner_join(households_2,households_3,by="ID")
colnames(households_2_3) <- c("Index2","ID","Index3")

ggplot(data = households_2_3) + 
  geom_point(aes(x = Index2,y = Index3), color = "royalblue4", 
             alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_smooth(aes(x = Index2,y = Index3),
              color = "red2", method = "lm") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Household Wealth Scores") +
  xlab("Index 2") +
  ylab("Index 3") +
  xlim(-0.8,2.15) +
  ylim(-0.8,2.15) +
  coord_equal(ratio = 1) 
```

Index 2 vs Index 4:
```{r scatter 2vs4, echo = FALSE}
households_2_4 <- inner_join(households_2,households_4,by="ID")
colnames(households_2_4) <- c("Index2","ID","Index4")

ggplot(data = households_2_4) + 
  geom_point(aes(x = Index2,y = Index4), color = "royalblue4", 
             alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_smooth(aes(x = Index2,y = Index4), 
              color = "red2", method = "lm") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Household Wealth Scores") +
  xlab("Index 2") +
  ylab("Index 4") +
  xlim(-0.8,2.15) +
  ylim(-0.8,2.15) +
  coord_equal(ratio = 1)
```

#####Household wealth scores of Index 1 vs Index 3
How does including appliance information affect the wealth index?
```{r scatter 1vs3, echo = FALSE}
households_1_3 <- inner_join(households_1,households_3,by="ID")
colnames(households_1_3) <- c("Index1","ID","Index3")

ggplot(data = households_1_3) + 
  geom_point(aes(x = Index1,y = Index3), color = "royalblue4", 
             alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_smooth(aes(x = Index1,y = Index3),
              color = "red2", method = "lm") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Household Wealth Scores") +
  xlab("Index 1") +
  ylab("Index 3") +
  xlim(-0.8,2.15) +
  ylim(-0.8,2.15) +
  coord_equal(ratio = 1) 
```