---
title: "Testing Number of People Per Room - Bin Cutoffs"
output:
  html_document: default
  html_notebook: default
---

####Note for user regarding this R Notebook:
This notebook explains our decision regarding the variable "number of people per room." Because this variable had a disproportionate number of levels compared to the other variables included in the MCA models, we binned the ranges 0-2 people per room, 3-9 people per room, and 10+ people per room, yielding three levels from more nearly 20. We found that both binning the variable and varying the bin cutoffs had little effect on the household wealth scores. In this section, we demonstrate these changes' effects with Indices 1 & 2.  

In order to knit the document fully, we must first include the code to build indices 1 & 2 (including loading and subsetting the data, re-leveling the variables, and building the MCA models), as the comparison plots reference the values generated by the MCA models. These pieces of code will make the notebook longer to knit, but will not appear in the output file.

####Pulling/manipulating data before importing to R
In IPUMS-I: Select variables to include as many of  
&nbsp;&nbsp;&nbsp;&nbsp;(COUNTRY, YEAR, SAMPLE, SERIAL, PERNUM, PERWT, RESIDENT preselected),  
&nbsp;&nbsp;&nbsp;&nbsp;HHWT, PERSONS, OWNERSHIP, ELECTRIC, WATSUP, SEWAGE, FUELCOOK, TRASH, HEAT, AIRCON, ROOMS, TOILET, FLOOR, WALL, ROOF,  
&nbsp;&nbsp;&nbsp;&nbsp;PHONE, CELL, AUTOS, REFRIG, TV, RADIO,  
&nbsp;&nbsp;&nbsp;&nbsp;CHBORN, SCHOOL, EDATTAIN, AND OCCISCO  
as are available.

In JMP: Uncheck "Use value labels" for all variables that were not preselected.

Note: Search for USER ACTION to find places in this code that require user consideration for each country sample. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(dplyr)
library(magrittr)
library(FactoMineR)
library(ggplot2)
```

```{r data source, include = FALSE}
#USER ACTION: replace "Zambia_2010.csv" with data from country of interest
#country <- read.csv("<csv filename here>")
country <- read.csv("Zambia_2010.csv")
```

```{r subsetting detailed, include = FALSE}
#Index 1
#USER ACTION: Comment out lines of code for variables unavailable in country sample:
country_1 = country %>% 
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier, 
  Ownership.of.dwelling..general.version., 
  Person.number, 
  Residence.status..de.facto..de.jure, 
  Children.ever.born, 
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general, 
  Telephone.availability,
  Cellular.phone.availability, 
  Automobiles.available, 
  Refrigerator, 
  Television.set, 
  Radio.in.household
  ))

#Index 2
country_2 = country %>% 
  filter(Person.number == 1) %>%
  select(-c(
  IPUMS.sample.identifier, 
  Ownership.of.dwelling..detailed.version., 
  Person.number, 
  Residence.status..de.facto..de.jure, 
  Children.ever.born, 
  School.attendance,
  Educational.attainment..international.recode..general.version.,
  Educational.attainment..international.recode..detailed.version.,
  Occupation..ISCO.general, 
  Trash.disposal,
  #Central.heating,
  #Air.conditioning,
  Telephone.availability,
  Cellular.phone.availability, 
  Automobiles.available, 
  Refrigerator, 
  Television.set, 
  Radio.in.household
  ))
```

```{r function not in, include = FALSE}
#Make "Not in" function (used for all indices)
'%!in%' <- function(x,y)!('%in%'(x,y))
```

```{r, include = FALSE}
# USER ACTION: Check this code for each data set. Comment out lines of code for variables unavailable in country sample.  
# Note: Pay special attention to SEWAGE (see notes below) - needs to be checked for each sample  
# Note: As written, this code removes NIU, Unknown for each variable. User may remove or change filter commands to include NIU, Unknown.
```

```{r releveling index1, include = FALSE}
#Making variables for Index 1

#Making ownership.1 (Owned, Rented, or Other ownership status)
country_1 <- country_1 %>% filter(Ownership.of.dwelling..detailed.version. %!in% c(0,999)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..detailed.version. %in% 100:194, "Owned", ifelse(Ownership.of.dwelling..detailed.version. %in% 210:239, "Rented", "Other ownership status")))

#Making electric.1 (Has electricity or No electricity)
country_1 <- country_1 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply.1 (Access to piped water or No access to piped water)
country_1 <- country_1 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Note: Sewage is a variable that needs manual attention for Indices 1 and 3. Either, Sewage is ready to be simplified for Index 1 or Sewage is in the form intended for Index 2. In the second case, Sewage may be excluded or included (in simplified form) in Index 1 (at this stage of the research). A more formal or educated recommendation may follow in the course of our research.
# Run <unique(country_1$Sewage)> to check form of SEWAGE for sample.

#Making sewage.simple.1 (Sewage access [septic or public] or No sewage disposal)
country_1 <- country_1 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.simple.1 = ifelse(Sewage == 20, "No sewage disposal", "Sewage access"))

# #Making sewage.detailed.1 (Sewage system, Septic system, or No sewage)
# country_1 <- country_1 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.detailed.1 = ifelse(Sewage == 11, "Sewage system", ifelse(Sewage == 12, "Septic system", "No sewage")))

#Making cooking.fuel.1 (Cook with electricity, Cook with fossil fuel, Cook with solid fuel, or Cook with other fuel)
country_1 <- country_1 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel == 20, "Cook with electricity", ifelse(Cooking.fuel %in% c(30:47,62:65), "Cook with fossil fuel", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel"))))

#Making trash.1 (Collected by service or No sanitation service)
country_1 <- country_1 %>% filter(Trash.disposal %!in% c(0,99)) %>% mutate(trash.1 = ifelse(Trash.disposal %in% 10:14, "Trash collected by service", "No sanitation service"))

# #Making heat.1 (Central heating or No central heating)
# country_1 <- country_1 %>% filter(Central.heating %!in% c(0,9)) %>% mutate(heat.1 = ifelse(Central.heating %in% c(1, 5:7), "No central heating", "Central heating"))
# 
# #Alternative: If available for country sample, make aircon.1 to fill in for heat (No air conditioning or Air conditioning)
# country_1 <- country_1 %>% filter(Air.conditioning %!in% c(0,99)) %>% mutate(aircon.1 = ifelse(Air.conditioning == 10, "No air conditioning", "Air conditioning"))

#Note: We kept Number.of.rooms as an integer, assuming that making all households with code 30 (30+ rooms) equate to 30 rooms would have a negligable effect on the index; very few houses will have greater than 30 rooms, and households with enough wealth for 30 rooms will be ranked similarly to those with greater than 30.
#Note: We changed Number.of.rooms code 00 (Part of a room; no rooms) to 1 in order to do the calculation (Number.of.person.records/Number.of.rooms), assuming that having part of a room or no room is similar enough to having only one room to have a negligable effect.
#Note: We rounded number.ppl.per.room.1 to limit the levels that would result in the MCA to only integer values instead of all possible numeric values
#Note: We also tried binning number.ppl.per.room.1 to be in ranges: 0-2, 3-9, 10+ people per room, as the initial MCA gave disproportional weight to the levels of number.ppl.per.room.1.

#Making number.ppl.per.room.1
country_1 <- country_1 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_1 <- country_1 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/number.of.rooms.1)
country_1$number.ppl.per.room.1 <- round(country_1$number.ppl.per.room.1)
country_1$number.ppl.per.room.1 <- as.integer(country_1$number.ppl.per.room.1) 

#Different cutoffs for binning: Making bin.number.ppl.per.room.1
country_1 <- country_1 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_1 <- country_1 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Flush toilet, Non-flush toilet, No toilet)
country_1 <- country_1 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet == 10, "No toilet",ifelse(Toilet == 21, "Flush toilet","Non-flush toilet")))

#Making floor.1 (Finished or Unfinished)
country_1 <- country_1 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), "Finished floors", "Unfinished floors"))

#Making wall.1 (Cement or Not cement)
country_1 <- country_1 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), "Cement walls", "Not cement walls"))

#Making roof.1 (Cement or Not cement)
country_1 <- country_1 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), "Cement roof", "Not cement roof"))
```

```{r releveling index2, include = FALSE}
#Making variables for Index 2

#Making ownership.1 (Owned or Not owned)
country_2 <- country_2 %>% filter(Ownership.of.dwelling..general.version. %!in% c(0,9)) %>% mutate(ownership.1 = ifelse(Ownership.of.dwelling..general.version. == 1, "Owned","Not owned"))

#Making electric.1 (Has electricity or No electricity)
country_2 <- country_2 %>% filter(Electricity %!in% c(0,9)) %>% mutate(electric.1 = ifelse(Electricity == 1, "Has electricity", "No electricity"))

#Making water.supply.1 (Access to piped water or No access to piped water)
country_2 <- country_2 %>% filter(Water.supply %!in% c(0,99)) %>% mutate(water.supply.1 = ifelse(Water.supply %in% 10:18, "Access to piped water","No access to piped water"))

#Making sewage.general.1 (Sewage system or No sewage system)
country_2 <- country_2 %>% filter(Sewage %!in% c(0,99)) %>% mutate(sewage.general.1 = ifelse(Sewage %in% 10:12, "Sewage system", "No sewage system"))

#Making cooking.fuel.1 (Cook with fossil fuel or elec, Cook with solid fuels, or Cook with other fuel)
country_2 <- country_2 %>% filter(Cooking.fuel %!in% c(0,99)) %>% mutate(cooking.fuel.1 = ifelse(Cooking.fuel %in% c(20,30:47,62:65), "Cook with fossil fuel or elec", ifelse(Cooking.fuel %in% c(50:56,61), "Cook with solid fuels", "Cook with other fuel")))

#See Index 1 number.ppl.per.room.1 for notes regarding this calculation
#Making number.ppl.per.room.1
country_2 <- country_2 %>% filter(Number.of.rooms %!in% c(98,99)) %>% mutate(Number.of.rooms.1 = ifelse(Number.of.rooms == 0, 1, Number.of.rooms))
country_2 <- country_2 %>% mutate(number.ppl.per.room.1 = Number.of.person.records.in.the.household/Number.of.rooms.1)
country_2$number.ppl.per.room.1 <- round(country_2$number.ppl.per.room.1)
country_2$number.ppl.per.room.1 <- as.integer(country_2$number.ppl.per.room.1) 

#Different binning: Making bin.number.ppl.per.room.1
country_2 <- country_2 %>% mutate(bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:2, "0-2 ppl per room",ifelse(number.ppl.per.room.1 %in% 3:9, "3-9 ppl per room","10+ ppl per room")))

#Further split cutoffs for binning: Making alt.bin.number.ppl.per.room.1
country_2 <- country_2 %>% mutate(alt.bin.number.ppl.per.room.1 = ifelse(number.ppl.per.room.1 %in% 0:1, "0-1 ppl per room", ifelse(number.ppl.per.room.1 %in% 2:4, "2-4 ppl per room", ifelse(number.ppl.per.room.1 %in% 5:9, "5-9 ppl per room", "10+ ppl per room"))))

#Making toilet.1 (Toilet or No toilet)
country_2 <- country_2 %>% filter(Toilet %!in% c(0,99)) %>% mutate(toilet.1 = ifelse(Toilet %in% 10:11, "No toilet","Toilet"))

#Making dwelling.1 (Has cement structures [cement floor, cement wall, or cement roof] or has No cement structures) - need to create each of floor.1, wall.1, roof.1 first
#floor.1 (1 [Finished] or 0 [Unfinished])
country_2 <- country_2 %>% filter(Floor.material %!in% c(0,999)) %>% mutate(floor.1 = ifelse(Floor.material %in% c(200:234,236), 1, 0))

#wall.1 (1 [Cement] or 0 [Not cement])
country_2 <- country_2 %>% filter(Wall.or.building.material %!in% c(0,999)) %>% mutate(wall.1 = ifelse(Wall.or.building.material %in% c(500:513,515:523,545,546), 1, 0))

#roof.1 (1 [Cement] or 0 [Not cement])
country_2 <- country_2 %>% filter(Roof.material %!in% c(0,99)) %>% mutate(roof.1 = ifelse(Roof.material %in% c(10:24,29), 1, 0))

#floor.wall.roof.1 (sum of floor.1, wall.1, roof.1)
country_2 <- country_2 %>% mutate(floor.wall.roof.1 = floor.1+wall.1+roof.1)
#dwelling.1
country_2 <- country_2 %>% mutate(dwelling.1 = ifelse(floor.wall.roof.1 > 0, "Cement structures", "No cement structures"))
```

```{r change to factors, include = FALSE}
# USER ACTION:
# Note: Comment out lines of code for variables unavailable in country sample.  
# Note: Review Sewage variable choice before running this chunk. 

#Index 1
country_1$ownership.1 = as.factor(country_1$ownership.1)
country_1$electric.1 = as.factor(country_1$electric.1)
country_1$water.supply.1 = as.factor(country_1$water.supply.1)

#Sewage: Comment out the irrelevant variable (sewage.<simple or detailed>.1)
country_1$sewage.simple.1 = as.factor(country_1$sewage.simple.1)
#country_1$sewage.detailed.1 = as.factor(country_1$sewage.detailed.1)

country_1$cooking.fuel.1 = as.factor(country_1$cooking.fuel.1)
country_1$trash.1 = as.factor(country_1$trash.1)
#country_1$heat.1 = as.factor(country_1$heat.1)
#country_1$aircon.1 = as.factor(country_1$aircon.1)
country_1$number.ppl.per.room.1 = as.factor(country_1$number.ppl.per.room.1)
country_1$bin.number.ppl.per.room.1 = as.factor(country_1$bin.number.ppl.per.room.1)
country_1$alt.bin.number.ppl.per.room.1 = as.factor(country_1$alt.bin.number.ppl.per.room.1)
country_1$toilet.1 = as.factor(country_1$toilet.1)
country_1$floor.1 = as.factor(country_1$floor.1)
country_1$wall.1 = as.factor(country_1$wall.1)
country_1$roof.1 = as.factor(country_1$roof.1)


#Index 2
country_2$ownership.1 = as.factor(country_2$ownership.1)
country_2$electric.1 = as.factor(country_2$electric.1)
country_2$water.supply.1 = as.factor(country_2$water.supply.1)
country_2$sewage.general.1 = as.factor(country_2$sewage.general.1)
country_2$cooking.fuel.1 = as.factor(country_2$cooking.fuel.1)
country_2$number.ppl.per.room.1 = as.factor(country_2$number.ppl.per.room.1)
country_2$bin.number.ppl.per.room.1 = as.factor(country_2$bin.number.ppl.per.room.1)
country_2$alt.bin.number.ppl.per.room.1 = as.factor(country_2$alt.bin.number.ppl.per.room.1)
country_2$toilet.1 = as.factor(country_2$toilet.1)
country_2$dwelling.1 = as.factor(country_2$dwelling.1)
```

```{r serial numbers, include = FALSE}
# USER ACTION: If using a sample containing more than one country, the row name needs to contain the full IPUMS idendifier and Household.serial.number to track the individual houses through the full MCA analysis. 

row.names(country_1) = country_1$Household.serial.number
row.names(country_2) = country_2$Household.serial.number
```

```{r mca1234, include = FALSE}
#USER ACTION: Comment out variables unavailable in the country sample:

country_1.sub = select(country_1, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.simple.1, 
                                      #sewage.detailed.1,
                                      cooking.fuel.1,
                                      trash.1,
                                      #heat.1,
                                      #aircon.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      floor.1,
                                      wall.1,
                                      roof.1))

country_2.sub = select(country_2, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.general.1,
                                      cooking.fuel.1,
                                      bin.number.ppl.per.room.1,
                                      toilet.1,
                                      dwelling.1))

country_1.mca = MCA(country_1.sub,row.w = country_1$Household.weight, graph = FALSE)
country_2.mca = MCA(country_2.sub,row.w = country_2$Household.weight, graph = FALSE)
```

```{r household score, include = FALSE}
#Index 1
households_1 <- data.frame(country_1.mca$ind$coord[,1])
households_1$ID <- rownames(households_1)
colnames(households_1) <- c("score","ID")

#Index 2
households_2 <- data.frame(country_2.mca$ind$coord[,1])
households_2$ID <- rownames(households_2)
colnames(households_2) <- c("score","ID")
```

##Binning Comparisons
USER ACTION: Begin reviewing code for binning comparisons here.

###Comparing number.ppl.per.room.1 binned (as in Index 1) vs unbinned (as pulled from IPUMS-I)

Building the MCA models using "number of people per room" as read from IPUMS-I (Index 1, 2 unbinned):
```{r appdx mca_unbinned}
country_1.sub.unbinned = select(country_1, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.simple.1, 
                                      #sewage.detailed.1,
                                      cooking.fuel.1,
                                      trash.1,
                                      #heat.1,
                                      #aircon.1,
                                      number.ppl.per.room.1,
                                      toilet.1,
                                      floor.1,
                                      wall.1,
                                      roof.1))

country_2.sub.unbinned = select(country_2, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.general.1,
                                      cooking.fuel.1,
                                      number.ppl.per.room.1,
                                      toilet.1,
                                      dwelling.1))

country_1.mca.unbinned = MCA(country_1.sub.unbinned,row.w = country_1$Household.weight, graph = FALSE)
country_2.mca.unbinned = MCA(country_2.sub.unbinned,row.w = country_2$Household.weight, graph = FALSE)
```

#####Comparing Loadings 
The plot of loadings for unbinned number.ppl.per.room.1 show how people per room has a disproportionate number of levels.  
Index 1 unbinned and Index 1:  
```{r appdx 1 unbinned loadings, echo = FALSE}
#Index 1, unbinned
loadings_1.unbinned = data.frame(country_1.mca.unbinned$var$coord[,1])
colnames(loadings_1.unbinned) = "weights"
loadings_1.unbinned$levels = rownames(loadings_1.unbinned)
ggplot(data = loadings_1.unbinned) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 1 Unbinned Loadings")+ 
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")

#Index 1
loadings_1 = data.frame(country_1.mca$var$coord[,1])
colnames(loadings_1) = "weights"
loadings_1$levels = rownames(loadings_1)
ggplot(data = loadings_1) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 1 Loadings") + 
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")
```

Index 2 unbinned and Index 2:  
```{r appdx 2 unbinned loadings, echo = FALSE}
#Index 2, unbinned
loadings_2.unbinned = data.frame(country_2.mca.unbinned$var$coord[,1])
colnames(loadings_2.unbinned) = "weights"
loadings_2.unbinned$levels = rownames(loadings_2.unbinned)
ggplot(data = loadings_2.unbinned) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 2 Unbinned Loadings") +
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")

#Index 2
loadings_2 = data.frame(country_2.mca$var$coord[,1])
colnames(loadings_2) = "weights"
loadings_2$levels = rownames(loadings_2)
ggplot(data = loadings_2) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 2 Loadings") +
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")
```

#####Comparing household scores  
Household score for Index 1, 2 unbinned:
```{r appdx household scores}
#Index 1 unbinned
households_1.unbinned <- data.frame(country_1.mca.unbinned$ind$coord[,1])
households_1.unbinned$ID <- rownames(households_1.unbinned)
colnames(households_1.unbinned) <- c("score","ID")

#Index 2 unbinned
households_2.unbinned <- data.frame(country_2.mca.unbinned$ind$coord[,1])
households_2.unbinned$ID <- rownames(households_2.unbinned)
colnames(households_2.unbinned) <- c("score","ID")
```

Plots of correlation between household score for Index 1 vs unbinned Index 1, Index 2 vs unbinned Index 2 demonstrate that the two binning schemes hardly differ. The red line represents what would be a perfect correlation.  
Index 1 vs Index 1 unbinned:  
```{r appdx plot b/ub ind1, echo = FALSE}
#Join indices' household scores
households_1_1ub <- inner_join(households_1,households_1.unbinned,by="ID")
colnames(households_1_1ub) <- c("Index1","ID","UBIndex1")

#Scatterplot of scores
ggplot(data = households_1_1ub) + 
  geom_point(aes(x = Index1,y = UBIndex1), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 1 vs Index 1 Unbinned") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 1") +
  ylab("Index 1, Unbinned")
```
Index 2 vs Index 2 unbinned:   
```{r appdx plot b/ub ind2, echo = FALSE}
#Join indices' household scores
households_2_2ub <- inner_join(households_2,households_2.unbinned,by="ID")
colnames(households_2_2ub) <- c("Index2","ID","UBIndex2")

#Scatterplot of scores
ggplot(data = households_2_2ub) + 
  geom_point(aes(x = Index2,y = UBIndex2), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 2 vs Index 2 Unbinned") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 2") +
  ylab("Index 2, Unbinned")
```

###Comparing number.ppl.per.room.1 binned (as in Index 1) vs alternate binning (0-1, 2-4, 5-9, 10+)

Building the MCA models using "number of people per room" as binned above (Index 1, 2 altbin):
```{r appdx mca_altbin}
country_1.sub.altbin = select(country_1, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.simple.1, 
                                      #sewage.detailed.1,
                                      cooking.fuel.1,
                                      trash.1,
                                      #heat.1,
                                      #aircon.1,
                                      alt.bin.number.ppl.per.room.1,
                                      toilet.1,
                                      floor.1,
                                      wall.1,
                                      roof.1))

country_2.sub.altbin = select(country_2, c(ownership.1, 
                                      electric.1, 
                                      water.supply.1, 
                                      sewage.general.1,
                                      cooking.fuel.1,
                                      alt.bin.number.ppl.per.room.1,
                                      toilet.1,
                                      dwelling.1))

country_1.mca.altbin = MCA(country_1.sub.altbin,row.w = country_1$Household.weight, graph = FALSE)
country_2.mca.altbin = MCA(country_2.sub.altbin,row.w = country_2$Household.weight, graph = FALSE)
```

#####Comparing Loadings 
The plot of loadings for alternate binning of number.ppl.per.room.1 show that the index weights the levels similarly to our selected binning.  
Index 1 altbin and Index 1:  
```{r appdx 1 altbin, echo = FALSE}
#Index 1, altbin
loadings_1.altbin = data.frame(country_1.mca.altbin$var$coord[,1])
colnames(loadings_1.altbin) = "weights"
loadings_1.altbin$levels = rownames(loadings_1.altbin)
ggplot(data = loadings_1.altbin) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 1 Altbin Loadings") + 
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")

#Index 1
loadings_1 = data.frame(country_1.mca$var$coord[,1])
colnames(loadings_1) = "weights"
loadings_1$levels = rownames(loadings_1)
ggplot(data = loadings_1) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 1 Loadings") + 
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")
```
Index 2 altbin and Index 2:  
```{r appdx 2 altbin, echo = FALSE}
#Index 2, altbin
loadings_2.altbin = data.frame(country_2.mca.altbin$var$coord[,1])
colnames(loadings_2.altbin) = "weights"
loadings_2.altbin$levels = rownames(loadings_2.altbin)
ggplot(data = loadings_2.altbin) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 2 Altbin Loadings") +
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")

#Index 2
loadings_2 = data.frame(country_2.mca$var$coord[,1])
colnames(loadings_2) = "weights"
loadings_2$levels = rownames(loadings_2)
ggplot(data = loadings_2) + 
  geom_bar(aes(x = reorder(levels,weights), y = weights), stat = 'identity', 
           col = "darkblue", fill = "darkblue", alpha = 0.4) +
  theme(axis.text.x = element_text(angle = 90), 
        plot.title = element_text(size = 20, face = "bold")) +
  ggtitle("Index 2 Loadings") +
  xlab("Dwelling Characteristic") +
  ylab("Index Weight")
```

###Comparing Household Scores
Household scores for Index 1, 2 altbin:  
```{r appdx household scores 2}
#Index 1, altbin
households_1.altbin <- data.frame(country_1.mca.altbin$ind$coord[,1])
households_1.altbin$ID <- rownames(households_1.altbin)
colnames(households_1.altbin) <- c("score","ID")

#Index 2, altbin
households_2.altbin <- data.frame(country_2.mca.altbin$ind$coord[,1])
households_2.altbin$ID <- rownames(households_2.altbin)
colnames(households_2.altbin) <- c("score","ID")
```

Plotting the correlation between the alternate binning of num.ppl.per.room.1 for Indices 1 and 2 vs the unbinned versions demonstrates that num.ppl.per.room.1 in Index 1 and 2 can be binned using either binning scheme with similar results.  
Plot of correlation between household score for binned (Index 1) vs Altbin Index 1:  
```{r appdx plot b/ab ind1, echo = FALSE}
#Join indices' household scores
households_1_1ab <- inner_join(households_1,households_1.altbin,by="ID")
colnames(households_1_1ab) <- c("Index1","ID","ABIndex1")

#Scatterplot of scores
ggplot(data = households_1_1ab) + 
  geom_point(aes(x = Index1,y = ABIndex1), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 1 vs Index 1 Altbin") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 1, Binned") +
  ylab("Index 1, Altbin")
```
Plot of correlation between household score for binned vs Altbin Index 2:  
```{r appdx plot b/ab ind2, echo = FALSE}
#Join indices' household scores
households_2_2ab <- inner_join(households_2,households_2.altbin,by="ID")
colnames(households_2_2ab) <- c("Index2","ID","ABIndex2")

#Scatterplot of scores
ggplot(data = households_2_2ab) + 
  geom_point(aes(x = Index2,y = ABIndex2), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 2 vs Index 2 Altbin") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 2, Binned") +
  ylab("Index 2, Altbin")
```

Plots of correlation between household score for Index 1 with alternate bining vs Index 1 unbinned, Index 2 with alternate binning vs Index 2 unbinned demonstrate that the two binning schemes hardly differ.  
Plot of correlation between household score for altbinned vs unbinned Index 1:  
```{r appdx plot households ub/ab ind1, echo = FALSE}
#Index 1 altbin, unbinned
households_1ab_1ub <- inner_join(households_1.altbin,households_1.unbinned,by="ID")
colnames(households_1ab_1ub) <- c("ABIndex1","ID","UBIndex1")

#Index 1 altbin, unbinned
ggplot(data = households_1ab_1ub) + 
  geom_point(aes(x = ABIndex1,y = UBIndex1), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 1 Altbin vs Index 1 Unbinned") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 1, Altbin") +
  ylab("Index 1, Unbinned")
```

Plot of correlation between household score for altbinned vs unbinned Index 2:  
```{r appdx plot households ub/ab ind2, echo = FALSE}
#Index 2 altbin, unbinned
households_2ab_2ub <- inner_join(households_2.altbin,households_2.unbinned,by="ID")
colnames(households_2ab_2ub) <- c("ABIndex2","ID","UBIndex2")

#Index 2 altbin, unbinned
ggplot(data = households_2ab_2ub) + 
  geom_point(aes(x = ABIndex2,y = UBIndex2), color = "royalblue4", 
              alpha = 0.25, shape = 21, fill = "royalblue2") +
  geom_abline(slope = 1,intercept=0, color = "Red") +
  ggtitle("Comparing Index 2 Altbin vs Index 2 Unbinned") +
  theme(plot.title = element_text(size = 20, face = "bold")) +
  xlab("Index 2, Altbin") +
  ylab("Index 2, Unbinned")
```

